Viscous Transport in Eroding Porous Media

### 2019-08-23 docs/transport/erosion_transport.tex
working on section 5

Given the discretization points of $\bd\Omega$, applying the trapezoid
rule results in the collocation method for~\eqref{eqn:BIE} 
\begin{subequations}
\label{eqn:trapLinearSystem}
  \begin{alignat}{3}
  \UU_\ell &= \sum_{j=1}^{N_\out} 
    w^0_j D(\yy^0_\ell,\yy^0_j) \eeta_j +
  \sum_{k=1}^M \sum_{j=1}^{N_\iin}
    w^k_j D(\yy^0_\ell,\yy^k_j) \eeta^k_j +
  \sum_{j=1}^{N_\out} w^0_j N_0(\yy^0_\ell,\yy^0_j) \nonumber \\
  &\quad+\sum_{k=1}^M S[\llambda_k,\cc_k](\yy^0_\ell) + 
  \sum_{k=1}^M R[\xi_k,\cc_k](\yy^0_\ell),
    &&\hspace{-30pt} \ell = 1,\ldots,N_\out, \\
%
  \mathbf{0} &= \sum_{j=1}^{N_\out} 
    w^0_j D(\yy^m_\ell,\yy^0_j) \eeta_j +
  \sum_{k=1}^M \sum_{j=1}^{N_\iin}
    w^k_j D(\yy^m_\ell,\yy^k_j) \eeta^k_j \nonumber \\
  &\quad+\sum_{k=1}^M S[\llambda_k,\cc_k](\yy^m_\ell) + 
  \sum_{k=1}^M R[\xi_k,\cc_k](\yy^m_\ell),
    &&\hspace{-30pt}\ell = 1,\ldots,N_\iin, \: m=1,\ldots,M, \\
%
  \llambda_m &= \frac{1}{2\pi} \sum_{j=1}^{N_\iin} \eeta^m_j,  \quad
  \xi_m = \frac{1}{2\pi} \sum_{j=1}^{N_\iin}
    (\yy^m_j - \cc_m)^\perp \cdot \eeta^m_j,
    &&\hspace{-30pt} m=1,\ldots,M,
\end{alignat}
\end{subequations}
where $w^k_j$ are quadrature weights (which depend on $N_\iin$,
$N_\out$, and the lengths of $\gamma^k$), and $D$ is the kernel of the
Stokes double-layer potential defined in
Equation~\eqref{eqn:velocityDLP}.  The linear
system~\eqref{eqn:trapLinearSystem} is a well-conditioned second-kind
integral equation that is solved iteratively with GMRES.  If the number
of discretization points is sufficiently large, then the solution
of~\eqref{eqn:trapLinearSystem} is an accurate approximation of the
density function, Stokeslets, and rotlets.   Then, if $\xx \in \Omega$
is sufficiently far from $\bd\Omega$, an approximation of the velocity,
deformation tensor, and vorticity due to the double-layer potential is
found by applying the trapezoid rule to \eqref{eqn:velocityDLP},
\eqref{eqn:deformationDLP}, and~\eqref{eqn:vorticityDLP}, respectively.
In particular,
\begin{align}
  \label{eqn:trap}
  \uu(\xx) &= \sum_{j=1}^{N_\out} w^0_j D(\xx,\yy^0_j) \eeta_j +
  \sum_{k=1}^M \sum_{j=1}^{N_\iin} w^k_j D(\xx,\yy^k_j) \eeta^k_j, \\
%
  \ssigma(\xx) &= \sum_{j=1}^{N_\out} w^0_j \mathbf{D}_\ssigma(\xx,\yy^0_j) \eeta_j +
  \sum_{k=1}^M \sum_{j=1}^{N_\iin} w^k_j \mathbf{D}_\ssigma(\xx,\yy^k_j)
  \eeta^k_j, \\
%
  w(\xx) &= \sum_{j=1}^{N_\out} w^0_j D_\omega(\xx,\yy^0_j) \eeta_j +
  \sum_{k=1}^M \sum_{j=1}^{N_\iin} w^k_j D_\omega(\xx,\yy^k_j)
  \eeta^k_j.
\end{align}
The contributions due to the Stokeslets and rotlets require no
quadrature and are easily included in the velocity, deformation tensor,
and vorticity.  Finally, to compute the shear stress, we apply the
trapezoid rule with Fourier differentiation to approximate the jump
term~\eqref{eqn:deformationJump} of the deformation tensor, and then the
tensor is applied to the normal and tangent vectors as defined in
equation~\eqref{eqn:shearStress}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Efficiently Applying the Quadrature}
\label{sec:fmm}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The Barycentric quadrature rule achieves acceptable accuracy for
computing the velocity, shear stress, and vorticity.  However, applied
directly, it requires $\mathcal{O}(N^2)$ operations.  By using a fast
summation method, such as the fast multipole method~\cite{gre-rok1987},
this cost can be reduced to $\mathcal{O}(N)$ operations.  However, the
Barycentric quadrature formulas require several $N$-body calculations,
and they must be done for each component of $\bd\Omega$ individually.
The resulting cost would be prohibitive, so we use a hybrid method that
combines the Barycentric quadrature rules and the trapezoid rule.  In
short, the fast multipole accelerated trapezoid rule is used to compute
the velocity at most of the discretization points, and the
non-accelerated Barycentric quadrature rule is used to locally correct
the trapezoid rule.

We start by using the trapezoid rule~\eqref{eqn:trap} accelerated with
the FMM, which requires $\mathcal{O}(N)$ operations, and we call this
velocity $\vv_\trap(\xx)$.  For target points that are sufficiently far
from $\bd\Omega$, $\vv_\trap(\xx)$ is sufficiently accurate.  However,
if $\xx$ is sufficiently close to one of the eroding bodies $\gamma_k$,
or the outer wall $\Gamma$, the velocity contribution from that
particular component of $\bd\Omega$ is inaccurate and it must be
corrected.  Assuming that $\xx$ is too close to $\gamma_k$, we first
subtract the velocity contribution of $\vv(\xx)$ due to $\gamma_k$.
Then, the Barycentric quadrature rule is used to compute the velocity
due to $\gamma_k$ with more accuracy.  That is, the final velocity is
\begin{align}
  \vv(\xx) = \vv_\trap(\xx) - \frac{L_k}{N_k} \sum_{j=1}^{N_\iin} 
    D(\xx,\yy^k_j) \eeta(\yy_j) + \vv^k_\bary(\xx),
  \label{eqn:velocityDecomp}
\end{align}
where $\vv^k_\bary(\xx)$ is the velocity at $\xx$ resulting from
applying the Barycentric quadrature rule due to $\gamma_k$.  This
strategy naturally extends to points that are close to $\Gamma$, and to
the scenario when $\xx$ is simultaneously close to multiple components
of $\bd\Omega$.  While the term $\vv_\trap(\xx)$ in
equation~\eqref{eqn:velocityDecomp} is computed for all target points
using the FMM, the other two terms are computed with a direct summation.
However, these terms are only required for target points near an eroding
body or the outer wall, and the number of such points is much less than
$N$.  An identical quadrature method is used to compute the vorticity
and shear stress.  That is, these quantities are computed using the
trapezoid rule, and then updates due to inaccuracies of the trapezoid
rule are made.  However, since the shear stress and vorticity are only
computed once per time step, the trapezoid rule is applied with a direct
summation rather than the FMM.  Relative to the cost of applying GMRES
with the FMM, the additional once-per-time-step cost is minimal.

Per grid point, evaluating the Barycentric quadrature rules dominates
the computational cost. Therefore, it is imperative that we only apply
the Barycentric quadrature when the trapezoid rule is inaccurate.  As a
rule of thumb, the trapezoid rule due to $\gamma_k$ is
accurate~\cite{bar2014} if
\begin{align}
  \label{eqn:TrapCutoff}
  d(\xx,\gamma_k) = \inf_{\yy \in \gamma_k} \|\xx - \yy\| > 
    5 \frac{L_k}{N_\iin}.
\end{align}
To avoid checking if all target point $\xx$
satisfy~\eqref{eqn:TrapCutoff}, we make use of the {\thL} variables.
First, given the two eroding bodies $\gamma_i$ and $\gamma_j$ with
lengths $L_i$ and $L_j$, and centers $\cc_i$ and $\cc_j$, we first check
if
\begin{align}
  \label{eqn:BodyCutoff}
  \|\cc_i - \cc_j\| < \frac{L_i}{2\pi} + \frac{L_j}{2\pi} + 
    \alpha_\iin \left(\frac{L_i}{N_\iin} + \frac{L_j}{N_\iin} \right),
\end{align}
where $\alpha_\iin \geq 1$ is a parameter that needs to be chosen.  In
this manner, rather than using an expensive all-to-all algorithm to
compute the distance between all pairs of discretization points, we
compute the distance between pairs of circles.  This criteria allows us
to quickly determine bodies that contain discretization points where the
Barycentric quadrature rule may need to be applied, and the parameter
$\alpha_\iin$ accounts for the fact that the bodies are being
approximated with circles.  Assuming that the two bodies $\gamma_i$ and
$\gamma_j$ satisfy condition~\eqref{eqn:BodyCutoff}, for each point $\xx
\in \gamma_j$, we check if
\begin{align}
  \label{eqn:PointsBodyCutoff}
  \|\xx - \cc_i\| < \frac{L_i}{2\pi}
+ \alpha_\iin \left(\frac{L_i}{N_\iin} \right).
\end{align}
To determine if points on $\gamma_i$ are too close to the outer wall, we
use the fact that the outer wall is bounded in $[-3,3] \times [-1,1]$,
and that the bodies are all contained in $[-1,1] \times [-1,1]$.  We
first check if
\begin{align}
  \left\|\cc_i - \left[
    \begin{array}{c}
      x_i \\ \pm 1
    \end{array}
    \right]
  \right\| < \frac{L_i}{2 \pi} + \alpha_\out \frac{L_\out}{N_\out}
\end{align}
where $x_i$ is the x coordinate of $\cc_i$. If body $\gamma_i$ satisfies
this condition, for each point $\xx=(x,y) \in \gamma_i$, we apply the
Barycentric rule to points that satisfy
\begin{align}
  \label{eqn:PointsWallCutoff}
  |y \pm 1| < \frac{L_i}{2 \pi} +\alpha_\out \frac{L_\out}{N_\out}.
\end{align}
Finally, to determine if a target point $\xx$ that is in the fluid bulk
rather than on another eroding body requires the Barycentric quadrature
rule, we only check conditions~\eqref{eqn:PointsBodyCutoff}
and~\eqref{eqn:PointsWallCutoff}.

>>>>

Given the discretization points of $\bd\Omega$, applying the trapezoid
rule results in the collocation method for~\eqref{eqn:BIE} 
\begin{subequations}
\label{eqn:trapLinearSystem}
  \begin{alignat}{3}
  \UU_\ell &= \sum_{j=1}^{N_\out} 
    w^0_j D(\yy^0_\ell,\yy^0_j) \eeta_j +
  \sum_{k=1}^M \sum_{j=1}^{N_\iin}
    w^k_j D(\yy^0_\ell,\yy^k_j) \eeta^k_j +
  \sum_{j=1}^{N_\out} w^0_j N_0(\yy^0_\ell,\yy^0_j) \nonumber \\
  &\quad+\sum_{k=1}^M S[\llambda_k,\cc_k](\yy^0_\ell) + 
  \sum_{k=1}^M R[\xi_k,\cc_k](\yy^0_\ell),
    &&\hspace{-60pt} \ell = 1,\ldots,N_\out, \\
%
  \mathbf{0} &= \sum_{j=1}^{N_\out} 
    w^0_j D(\yy^m_\ell,\yy^0_j) \eeta_j +
  \sum_{k=1}^M \sum_{j=1}^{N_\iin}
    w^k_j D(\yy^m_\ell,\yy^k_j) \eeta^k_j \nonumber \\
  &\quad+\sum_{k=1}^M S[\llambda_k,\cc_k](\yy^m_\ell) + 
  \sum_{k=1}^M R[\xi_k,\cc_k](\yy^m_\ell),
    &&\hspace{-60pt}\ell = 1,\ldots,N_\iin, \: m=1,\ldots,M, \\
%
  \llambda_m &= \frac{1}{2\pi} \sum_{j=1}^{N_\iin} \eeta^m_j,  \quad
  \xi_m = \frac{1}{2\pi} \sum_{j=1}^{N_\iin}
    (\yy^m_j - \cc_m)^\perp \cdot \eeta^m_j,
    &&\hspace{-60pt} m=1,\ldots,M,
\end{alignat}
\end{subequations}
where $w^k_j$ are quadrature weights (which depend on $N_\iin$,
$N_\out$, and the lengths of $\gamma^k$), and $D$ is the kernel of the
Stokes double-layer potential defined in
Equation~\eqref{eqn:velocityDLP}.  The linear
system~\eqref{eqn:trapLinearSystem} is a well-conditioned second-kind
integral equation that is solved iteratively with GMRES.  If the number
of discretization points is sufficiently large, then the solution
of~\eqref{eqn:trapLinearSystem} is an accurate approximation of the
density function, Stokeslets, and rotlets.   Then, if $\xx \in \Omega$
is sufficiently far from $\bd\Omega$, an approximation of the velocity,
deformation tensor, and vorticity due to the double-layer potential is
found by applying the trapezoid rule to \eqref{eqn:velocityDLP},
\eqref{eqn:deformationDLP}, and~\eqref{eqn:vorticityDLP}, respectively.
In particular,
\begin{align}
  \label{eqn:trap}
  \uu(\xx) &= \sum_{j=1}^{N_\out} w^0_j D(\xx,\yy^0_j) \eeta_j +
  \sum_{k=1}^M \sum_{j=1}^{N_\iin} w^k_j D(\xx,\yy^k_j) \eeta^k_j, \\
%
  \ssigma(\xx) &= \sum_{j=1}^{N_\out} w^0_j \mathbf{D}_\ssigma(\xx,\yy^0_j) \eeta_j +
  \sum_{k=1}^M \sum_{j=1}^{N_\iin} w^k_j \mathbf{D}_\ssigma(\xx,\yy^k_j)
  \eeta^k_j, \\
%
  w(\xx) &= \sum_{j=1}^{N_\out} w^0_j D_\omega(\xx,\yy^0_j) \eeta_j +
  \sum_{k=1}^M \sum_{j=1}^{N_\iin} w^k_j D_\omega(\xx,\yy^k_j)
  \eeta^k_j.
\end{align}
The contributions due to the Stokeslets and rotlets require no
quadrature and are easily included in the velocity, deformation tensor,
and vorticity.  Finally, to compute the shear stress, we apply the
trapezoid rule with Fourier differentiation to approximate the jump
term~\eqref{eqn:deformationJump} of the deformation tensor, and then the
tensor is applied to the normal and tangent vectors as defined in
equation~\eqref{eqn:shearStress}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Efficiently Applying the Quadrature}
\label{sec:fmm}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The Barycentric quadrature rule achieves acceptable accuracy for
computing the velocity, shear stress, and vorticity.  However, applied
directly, it requires $\mathcal{O}(N^2)$ operations.  By using a fast
summation method, such as the fast multipole method~\cite{gre-rok1987},
this cost can be reduced to $\mathcal{O}(N)$ operations.  However, the
Barycentric quadrature formulas require several $N$-body calculations,
and they must be done for each component of $\bd\Omega$ individually.
The resulting cost would be prohibitive, so we use a hybrid method that
combines the Barycentric quadrature rules and the trapezoid rule.  In
short, the fast multipole accelerated trapezoid rule is used to compute
the velocity at most of the discretization points, and the
non-accelerated Barycentric quadrature rule is used to locally correct
the trapezoid rule.  Note that for the source points of the layer
potential is always one of the eroding bodies or the outer wall, but the
target point can either be on another component of $\bd\Omega$ or it can
be in the fluid bulk $\Omega$.

To compute the velocity double-layer potential, we start by applying the
trapezoid rule~\eqref{eqn:trap} accelerated with the FMM.  This
calculation requires $\mathcal{O}(N)$ operations, and we call the
resulting velocity $\vv_\trap(\xx)$.  Depending on the number of
discretization points $N_\iin$ and $N_\out$, there is a region separated
from $\bd\Omega$ where the error of $\vv_\trap(\xx)$ is less than a
user-specified tolerance.  However, for the other points in $\Omega$,
the trapezoid rule needs to be replaced with a more accurate quadrature
rule.  Note, however, that since a point is typically only close to one
or two components of $\bd\Omega$, only the contribution of these nearby
bodies need to be replaced.  Assuming that $\xx$ is too close to
$\gamma_k$, we first subtract the inaccurate approximation of the
double-layer potential due to $\gamma_k$.  Then, the Barycentric
quadrature rule is used to compute the velocity due to $\gamma_k$ with
more accuracy.  That is, the final velocity is
\begin{align}
  \vv(\xx) = \vv_\trap(\xx) - \frac{L_k}{N_k} \sum_{j=1}^{N_\iin} 
    D(\xx,\yy^k_j) \eeta(\yy_j) + \vv^k_\bary(\xx),
  \label{eqn:velocityDecomp}
\end{align}
where $\vv^k_\bary(\xx)$ is the velocity at $\xx$ resulting from
applying the Barycentric quadrature rule to the double-layer potential
due to $\gamma_k$.  This strategy naturally extends to points that are
close to $\Gamma$, and to the scenario when $\xx$ is simultaneously
close to multiple components of $\bd\Omega$.  While the term
$\vv_\trap(\xx)$ in equation~\eqref{eqn:velocityDecomp} is computed for
all target points using the FMM, the other two terms are computed with a
direct summation.  However, these terms are only required for target
points near an eroding body or the outer wall, and these points make up
only a small fraction of the total number of points.

An identical strategy is used to compute the vorticity and the
deformation tensor.  That is, the trapezoid rule is used as a first pass
to form the vorticity and deformation tensor, and then local corrections
are made to replace the inaccuracies of the trapezoid rule.  However,
since the shear stress and vorticity are only computed once per time
step, the trapezoid rule is applied with a direct summation rather than
the FMM.  Relative to the cost of computing the velocity at each
iteration of GMRES with the FMM, the additional once-per-time-step cost
to compute the vorticity and deformation tensor is minimal.

Per grid point, applying the Barycentric quadrature rules dominates the
computational cost. Therefore, it is imperative that we only apply the
Barycentric quadrature when the trapezoid rule is inaccurate.  As a rule
of thumb, the trapezoid rule due to $\gamma_k$ achieves machine epsilon
accuracy if~\cite{bar2014}
\begin{align}
  \label{eqn:TrapCutoff}
  d(\xx,\gamma_k) = \inf_{\yy \in \gamma_k} \|\xx - \yy\| > 
    5 \frac{L_k}{N_\iin}.
\end{align}
To avoid checking if all target point $\xx$
satisfy~\eqref{eqn:TrapCutoff}, we make use of the {\thL} variables.
First, given the two eroding bodies $\gamma_i$ and $\gamma_j$ with
lengths $L_i$ and $L_j$, and centers $\cc_i$ and $\cc_j$, we check if
\begin{align}
  \label{eqn:BodyCutoff}
  \|\cc_i - \cc_j\| < \frac{L_i}{2\pi} + \frac{L_j}{2\pi} + 
    \alpha_\iin \left(\frac{L_i}{N_\iin} + \frac{L_j}{N_\iin} \right),
\end{align}
where $\alpha_\iin \geq 1$ is a parameter that needs to be chosen.  In
this manner, rather than using an expensive all-to-all algorithm to
compute the distance between all pairs of discretization points, we
compute the distance between pairs of circles.  This criteria allows us
to quickly determine bodies that contain discretization points where the
Barycentric quadrature rule may need to be applied, and the parameter
$\alpha_\iin$ accounts for the fact that the bodies are being
approximated with circles.  Assuming that the two bodies $\gamma_i$ and
$\gamma_j$ satisfy condition~\eqref{eqn:BodyCutoff}, for each point $\xx
\in \gamma_j$, we check if
\begin{align}
  \label{eqn:PointsBodyCutoff}
  \|\xx - \cc_i\| < \frac{L_i}{2\pi}
+ \alpha_\iin \left(\frac{L_i}{N_\iin} \right).
\end{align}
To determine if points on $\gamma_i$ are too close to the outer wall, we
recall that the outer wall is bounded in $[-3,3] \times [-1,1]$, and
that the bodies are all contained in $[-1,1] \times [-1,1]$.  Therefore,
a target point can only be too close to the lines $y = \pm 1$, so we
first check if
\begin{align}
  \left\|\cc_i - \left[
    \begin{array}{c}
      x_i \\ \pm 1
    \end{array}
    \right]
  \right\| < \frac{L_i}{2 \pi} + \alpha_\out \frac{L_\out}{N_\out},
\end{align}
\todo[inline]{Bryan is here}
where $x_i$ is the x coordinate of $\cc_i$. If body $\gamma_i$ satisfies
this condition, for each point $\xx=(x,y) \in \gamma_i$, we apply the
Barycentric rule to points that satisfy
\begin{align}
  \label{eqn:PointsWallCutoff}
  |y \pm 1| < \frac{L_i}{2 \pi} +\alpha_\out \frac{L_\out}{N_\out}.
\end{align}
Finally, to determine if a target point $\xx$ that is in the fluid bulk
rather than on another eroding body requires the Barycentric quadrature
rule, we only check conditions~\eqref{eqn:PointsBodyCutoff}
and~\eqref{eqn:PointsWallCutoff}.
